Scalable and Modular Verified Compilation via Automated Proofs for Declarative Transformations.

### 2025-01-31 paper.tex
final refinement pass

A fully end-to-end verified \emph{mainstream} compiler with a small
trusted code base remains somewhat of an unattainable holy grail.
End-to-end verified compilers do exist (e.g., CompCert~\cite{leroyCompCertFormallyVerified} and CakeML~\cite{kumarCakeMLVerifiedImplementation2014}), but the extra
verification burden makes compiler development challenging, relegating
such compilers to specific high-assurance niches. On the other hand,
there are tools which can fully automatically prove the correctness of a
specific compiler transformation (e.g., Alive~\cite{lopesAlive2BoundedTranslation2021}, which has gained
widespread adoption in the LLVM community), but such tools
often rely on SMT solvers to achieve their push-button automation,
making them unsuited in our quest for a minimal trusted code base. I
wish to investigate a middle road: can we verify the core transformation
algorithm separately from specific transformation instances, and then
develop proof automation for fully automatically showing local
correctness of a transformation in a way that composes into a proof of
global correctness?

Specifically, I'll develop two core components of a verified compiler:
an optimization pass based on peephole optimizations and a code
generator that lowers the compiler IR (Intermediate Representation)
produced by the optimizer into assembly. Of course, a fully end-to-end
compiler would need many more components, such as a verified front-end
that parses source code and transforms it into semantically equivalent
compiler IR, however such components will be out-of-scope for my PhD.

@@

For the optimization pass, we choose to operate on a compiler IR with the same semantics as LLVM.
Although others have formalized the semantics of LLVM IR in a proof assistant before
\cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
included no proof automation. This will to our knowledge be the first
effort to provide fully automated proofs for correctness of LLVM
optimizations in a way that doesn't have to trust an external solver (as
Alive~\cite{lopesAlive2BoundedTranslation2021} and even AliveInLean~\cite{leeAliveInLeanVerifiedLLVM2019} do). Removing
this dependency is important as SMT solvers are large, complex pieces of
software with a history of bugs~\cite{brummayerFuzzingDeltadebuggingSMT2009, mansurDetectingCriticalBugs2020}.

@@

After optimization, the next phase will be to generate assembly from our
optimized IR and, crucially, to verify that the generated assembly has
the same semantics as the optimized IR. For maximum trust, we'd like to
use authoritative models of the ISA semantics of whichever platform we
are compiling for to give semantics to the assembly program. Sail
(\cite{armstrongISASemanticsARMv8a2019}) provides such models: the
Sail model for Arm-A is automatically derived from the official ASL
reference, and their RISC-V model has been adopted by the RISC-V
Foundation. We'll phrase the code generator similarly: by having
fragments of compiler IR to match for, and corresponding fragments of
assembly to rewrite them with. Still, code generation presents some
unique challenges. Firstly, although the transformations themselves
should be simpler, the Sail semantic model will be much larger than our
handwritten LLVM semantics. Secondly, we have to deal with semantics of
different languages (the compiler IR and the assembly language). It is
likely that proof automation for code generation will thus need a
slightly different design than proof automation for peephole
optimizations.

Finally, to model the semantics of LLVM IR in a modular way, we'd like
to follow Zakowski et
al.~(\cite{zakowskiModularCompositionalExecutable2021}) and use
interaction trees (\cite{xiaInteractionTreesRepresenting2020}).
Unfortunately, interaction trees are \emph{coinductive} in nature, to
capture possibly non-terminating programs, and Lean currently lacks
support for coinductive type. Therefore, I'd like to continue the work
that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
work by Avigad et al.~\cite{avigadDataTypesQuotients2019}, and
continue development of a Lean library for coinductive types.

@@

Finally, we ask \textsc{RQ3} because we need interaction trees to define the semantics
of potentially infinite loops. 
We'll consider our interaction tree library successful if
we can provide a similar basic API, and in particular, if we can define such semantics
without referring to details of the underlying QPF construction directly.

@@

Note that the benefits of a verified compilation pipeline go far beyond
simply reducing the chance of bugs. We believe it will lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.

@@

\subsection{Lean-MLIR Today}\label{lean-mlir-today}

Lean-mlir is a framework for modelling the \emph{semantics} of MLIR
dialects in Lean, developed by my research group~\cite{bhatVerifyingPeepholeRewriting2024}, focusing on the
\emph{pure} (i.e., side-effect-free) fragment of MLIR dialects.

In particular, lean-mlir currently has a model of just the bitvector
arithmetic operations of LLVM IR, and a verified peephole rewriter that
can apply a pure peephole rewrite (i.e., neither the program fragment
being matched on nor the fragment being rewritten to contain
side-effecting operations) and, assuming the local correctness of every
rewrite, we've shown the rewriter preserves the global semantics.

Furthermore, the Lean FRO has, in collaboration with my research group
and me, been developing proof automation for deciding the equality
of fixed-width bitvector expressions\cite{bovingTamingBitvectorBestiary}. In
particular, this proof automation allows us to fully automatically
decide the local correctness of \emph{pure} optimizations. Note that
although this tactic employs an external (certificate-producing) SAT
solver, it also includes a verified certificate checker, so that the
final proofs do \emph{not} rely on the correctness of the solver. That
said, the checker would be too slow if ran inside the Lean kernel, so to
get decent performance, this tactic \emph{does} rely on the previously
mentioned feature to run the compiled version of a Lean tactic, thus expanding
the trusted code base to include the Lean compiler.
As the paper argues, this is still a much smaller trusted code base than a full
SMT solver.

@@

My contributions to lean-mlir are primarily focussed on expanding the framework to 
include support for impure operations with side effects. 
I wish to continue this effort by modelling all operations of the
dialects mentioned before, in particular the dialect dealing with memory.
The bitvector arithmetic fragment of the previous section is basically the same as \texttt{arith}, and the
framework also already models part of \texttt{scf} (in particular, it
models counting for-loops, but not while-loops with arbitrary conditions
yet), so the main missing piece is a model of memory and memory
operations.

The main innovation in this project will be to build proof automation in
Lean for translation validation on this compiler IR. That is, given a
source and target program (potentially including holes), the tool will
automatically decide the correctness of the transformation. To do this
in the presence of side effects, I'll build a symbolic simulator.

@@

We can broadly categorize earlier efforts to apply formal methods to
LLVM into those that formalize the semantics of LLVM, but perform manual
proofs (e.g., VeLLVM~\cite{zhaoFormalVerificationSSABased, zhaoFormalizingLLVMIntermediate2012} 
or work by Zakowski et al.~\cite{zakowskiModularCompositionalExecutable2021}) and tools
like Alive (\cite{lopesAlive2BoundedTranslation2021}) that perform
fully automated translation validation, but rely on SMT solvers.

@@

\section{Scaling High-Assurance Translation Validation to
Authoritative ISA
Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}

@@

By building proof automation in Lean that can automatically verify the correctness of
such lowering, I'll answer research question \textsc{RQ2}.

@@

Lopoukhine et al have developed a set of MLIR
dialects that together model RISC-V assembly~\cite{lopoukhineMultiLevelCompilerBackend}. Representing our target
for code generation as an MLIR dialect allows us to use Lean-MLIR,
enabling reuse of the core datastructures provided by the framework, but
the benefits extend much further! Namely, depending on which of the
dialects provided by Lopoukhine et al.~we use, our representation of
assembly can (a) be either register allocated, or as an assembly-like
SSA IR, and (b) use either structured (for/while loops) or unstructured
control flow (branches).

We'll use these progressively lower level IRs as targets for our code
generator, explicitly separating the instruction selection pass, which
we'll implement via a simple rewriter, from a register allocation pass.
Finally, recall that we chose to stick to structured control flow in the
optimization pass to preserve structure. We'll initially do the same in the code
generator, postponing the lowering from structured into unstructured
control flow to the last moment.

@@

Of course, the Sail semantics will operate on traditional assembly, with
registers allocated and unstructured branches. This should not be a
problem, as lowering a single instruction at any of the progressive
levels of abstraction down to regular assembly should not be difficult.

@@

Again, the core contribution will be proof automation. By treating
instruction selection as a sequence of rewrites, we merely need to
verify that each of these rewrites individually preserves semantics.

This seems like much the same problem that we tackled in the previous
chapters, but there are some subtleties. For one, the rewrites
themselves should be simpler, with many rewrites having only a single
instruction as source and target. However, a Sail-derived semantics
model will be much larger than the handwritten model we'll be using for
the compiler IR, necessitating many more reasoning steps to simplify the
semantics for a single instruction. Many of the details that make the
Sail model so large might be irrelevant in our context, so I'll
investigate ways to simplify the model under specific assumptions. For
example, by assuming we're only compiling user-level programs, we might
be able to simplify away any complexity to do with kernel-level
instructions in the model.

@@

During my internship at Amazon, I developed a symbolic simulator for a
simplified fragment of Arm assembly programs, using a handwritten model
of the ARM ISA semantics. My internship focused primarily on making this
symbolic simulator scale to being able to simulate hundreds of
instructions within a reasonable timeframe. I'll be using this symbolic
simulator as a basis for the translation validation tool I aim to
develop, adapting it to the slightly different design constraints in
this context.

@@

Islaris is a Coq development, which builds upon Isla and Iris separation logic to 
prove properties about assembly programs using the power and expressivity of a 
theorem prover~\cite{sammlerIslarisVerificationMachine2022}. Their strategy starts by using Isla to transform a program into a trace, and then proving properties about this trace.
To remove the need to trust the SMT solver from, Sammler et al also investigated
the use of automated translation validation, which certifies that the trace produced
by Isla agrees with the lowering of the Sail semantics to Coq. This approach 
worked well for RISC-V, but unfortunately did not scale to the much larger Arm model.

@@

In previous chapters, we've glossed over how, exactly, I'll give
semantics to the side-effectful version of our compiler IR. Zakowski et
al.~have written about the benefits of an interaction tree (ITree)
based semantics~\cite{zakowskiModularCompositionalExecutable2021, xiaInteractionTreesRepresenting2020}, so I'd like to use
ITrees also. Sadly, ITrees are coinductive, to properly capture infinite loops, 
and Lean does not support coinductive types. 
I'd like to address this, and build an ergonomic Lean library for ITrees,
answering research question \textsc{RQ3} in the process.

@@

Luckily, we can encode coinductives in Lean's existing logic, using
so-called \emph{quotients of polynomial functors} (QPFs). Avigad et
al.~have written about the low-level construction~\cite{avigadDataTypesQuotients2019}, 
upon which I expanded in my Master's thesis by prototyping a higher-level framework for defining
coinductive types, called QPFTypes. This framework is far from perfect,
and I'll spend some time fixing bugs in the framework. Still, the main
goal will be simply to define interaction trees; not to perfectly polish
the framework.

@@

The QPFTypes framework currently only has facilities for defining types;
to define corecursive functions that operate over ITrees we only have
the low-level corecursion principle to work with. However, in their
original development of ITrees, Xia et al. in fact provide a set of three
combinators which they claim make working with ITrees more tractable,
and remove the need for users to work directly with Coq's implementation
of corecursive function (which they claim is not compositional). As it
turns out, these combinators are rather similar to the corecursion
principle we do have available. So, for our ITree library
in Lean, we don't have to worry about general corecursive functions; it
suffices to implement the three basic combinators that Xia et
al.~provide.

@@

Isabelle, in particular, is relevant because it's construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

@@

@@

\subsection*{Scaling High-Assurance Translation Validation to
Authoritative ISA
Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}

>>>>

A fully end-to-end verified \emph{mainstream} compiler with a small
trusted code base remains somewhat of an unattainable holy grail.
End-to-end verified compilers do exist (e.g., CompCert~\cite{leroyCompCertFormallyVerified} and CakeML~\cite{kumarCakeMLVerifiedImplementation2014}), but the extra
verification burden makes compiler development challenging, relegating
such compilers to specific high-assurance niches. On the other hand,
there are tools which can fully automatically prove the correctness of a
specific compiler transformation (e.g., Alive~\cite{lopesAlive2BoundedTranslation2021}, which has gained
widespread adoption in the LLVM community), but such tools
often rely on SMT solvers to achieve their push-button automation,
making them unsuited in our quest for a minimal trusted code base. I
wish to investigate a middle road: can we verify the core transformation
algorithm separately from specific transformation instances, and then
develop proof automation to show local
correctness of a transformation, without needing input from a formal methods export?

I'll investigate two core components of a verified compiler:
an optimization pass based on peephole optimizations and a code
generator that lowers the compiler IR (Intermediate Representation)
produced by the optimizer into assembly. Of course, a fully end-to-end
compiler would need many more components, such as a verified front-end
that parses source code and transforms it into semantically equivalent
compiler IR, however such components will be out-of-scope for my PhD.

@@

For the optimization pass, we choose to operate on a compiler IR with the same semantics as LLVM.
Although others have formalized the semantics of LLVM IR in a proof assistant before
\cite{zhaoFormalVerificationSSABased, zakowskiModularCompositionalExecutable2021}, those efforts
included no proof automation. This will, to my knowledge, be the first
effort to provide fully automated proofs for correctness of LLVM
optimizations in a way that doesn't have to trust an external solver (as
Alive~\cite{lopesAlive2BoundedTranslation2021} and even AliveInLean~\cite{leeAliveInLeanVerifiedLLVM2019} do). Removing
this dependency is important as SMT solvers are large, complex pieces of
software with a history of bugs~\cite{brummayerFuzzingDeltadebuggingSMT2009, mansurDetectingCriticalBugs2020}.

@@

After optimization, the next phase will be to generate assembly from our
optimized IR and, crucially, to verify that the generated assembly has
compatible semantics. For maximum trust, we'd like to
use authoritative models of the ISA semantics of whichever platform we
are compiling for to give semantics to the assembly program. Sail~\cite{armstrongISASemanticsARMv8a2019} provides such models: the
Sail model for Arm-A is automatically derived from the official ASL
reference, and the RISC-V model has been adopted by the RISC-V
Foundation. We'll phrase the code generator similarly: by having
fragments of compiler IR to match for, and corresponding fragments of
assembly to rewrite them with. Still, code generation presents some
unique challenges. Firstly, although the transformations themselves
should be simpler, the Sail semantic model will be much larger than our
handwritten LLVM semantics. Secondly, we have to deal with semantics of
different languages (the compiler IR and the assembly language). It is
likely that proof automation for code generation will thus need a
slightly different design than proof automation for peephole
optimizations.

Finally, to model the semantics of LLVM IR in a modular way, we'd like
to follow Zakowski et
al.~(\cite{zakowskiModularCompositionalExecutable2021}) and use
interaction trees (\cite{xiaInteractionTreesRepresenting2020}).
Unfortunately, interaction trees are \emph{coinductive} in nature, to
capture possibly non-terminating programs, and Lean currently lacks
support for coinductive type. Therefore, I'd like to continue the work
that I started in my own master's thesis~\cite{keizerImplementingDefinitionalCodatatype}, which is based on earlier
work by Avigad\etal{}~\cite{avigadDataTypesQuotients2019}, and
continue development of a Lean library for coinductive types.

@@

Finally, I ask \textsc{RQ3} because we need interaction trees to define the semantics
of potentially infinite loops. 
We'll consider our interaction tree library successful if
we can provide a similar basic API, and in particular, if we can define the 
semantics we need without referring to details of the underlying QPF construction directly.

@@

Note that the benefits of a verified compilation pipeline go far beyond
simply reducing the chance of bugs. It could lead to a culture
shift in the way software is written. Traditionally, writing compiler
optimizations was a domain reserved for the wise elders who we could
trust to perform the appropriate ``black magic''. Reasoning about the
correctness of a transformation was simply too hard, so we had to rely
on interpersonal trust. Tools like Alive have already caused a shift in
this approach: nowadays PRs that propose adding a new peephole
optimization to LLVM can include a link to Alive's online interface
showing the correctness, meaning no further discussion or review is
required. However, this effect is still limited to making reviews for
proposed additions to LLVM easier.

@@

\subsection{Lean-MLIR Today}\label{Lean-MLIR-today}

Lean-MLIR is a framework for modelling the \emph{semantics} of MLIR
dialects in Lean, developed by my research group~\cite{bhatVerifyingPeepholeRewriting2024}, focusing on the
\emph{pure} (i.e., side-effect-free) fragment of MLIR dialects.

In particular, Lean-MLIR currently has a model of just the bitvector
arithmetic operations of LLVM IR, and a verified peephole rewriter that
can apply a pure peephole rewrite (i.e., neither the program fragment
being matched on nor the fragment being rewritten to contain
side-effecting operations) and, assuming the local correctness of every
rewrite, we've shown the rewriter preserves the global semantics.

Furthermore, the Lean FRO has, in collaboration with my research group
and me, been developing proof automation for deciding the equality
of fixed-width bitvector expressions~\cite{bovingTamingBitvectorBestiary}. In
particular, this proof automation allows us to fully automatically
decide the local correctness of \emph{pure} optimizations. Note that
although this tactic employs an external (certificate-producing) SAT
solver, it also includes a verified certificate checker, so that the
final proofs do \emph{not} rely on the correctness of the solver. That
said, the checker would be too slow if ran inside the Lean kernel, so to
get decent performance, this tactic \emph{does} rely on the previously
mentioned feature to run the compiled version of a Lean tactic, thus expanding
the trusted code base to include the Lean compiler.
As the paper argues, this is still a much smaller trusted code base than a full
SMT solver.

@@

My contributions to Lean-MLIR are primarily focussed on expanding the framework to 
include support for impure operations with side effects. 
I wish to continue this effort by modelling all operations of the
dialects mentioned before. In particular, the dialect dealing with memory.
The bitvector arithmetic fragment of the previous section is basically the same as \texttt{arith}, and the
framework also already models part of \texttt{scf} (in particular, it
models counting for-loops, but not while-loops with arbitrary conditions
yet), so the main missing piece is a model of memory and memory
operations.

The main innovation in this project will be to build proof automation in
Lean for translation validation on this compiler IR. That is, given a
source and target program (potentially including holes), the tool will
automatically decide the correctness of the transformation. To do this
in the presence of side effects, I'll build a symbolic simulator in Lean, 
which will rely on the previously mentioned bitvector solver to discharge pure
bitvector proof obligations.

@@

We can broadly categorize earlier efforts to apply formal methods to
LLVM into those that formalize the semantics of LLVM, but perform manual
proofs (e.g., VeLLVM~\cite{zhaoFormalVerificationSSABased, zhaoFormalizingLLVMIntermediate2012} 
or work by Zakowski\etal{}~\cite{zakowskiModularCompositionalExecutable2021}) and tools
like Alive~\cite{lopesAlive2BoundedTranslation2021} that perform
fully automated translation validation, but rely on SMT solvers.

@@

\section{Automatically Verified Instruction Selection using
Authoritative ISA Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models}

@@

By building proof automation in Lean that can automatically verify the correctness of
such lowerings, I'll answer research question \textsc{RQ2}.

@@

Lopoukhine\etal{}have developed a set of MLIR
dialects that together model RISC-V assembly~\cite{lopoukhineMultiLevelCompilerBackend}. Representing our target
for code generation as an MLIR dialect allows us to use Lean-MLIR,
enabling reuse of the core datastructures provided by the framework, but
the benefits extend much further! Namely, depending on which of the
dialects provided by Lopoukhine\etal{}~we use, our representation of
assembly can (a) be either register allocated, or an assembly-like
SSA IR, and (b) use either structured (for/while loops) or unstructured
control flow (branches).

We'll use these progressively lower level RISC-V IRs as targets for our code
generator, explicitly separating the instruction selection pass, which
we'll implement via a simple rewriter, from a register allocation pass.
Finally, recall that we chose to stick to structured control flow in the
optimization pass to preserve structure. We'll initially do the same in the code
generator, postponing the lowering from structured into unstructured
control flow to the last moment.

@@

Of course, the Sail semantics will operate on traditional assembly, with
registers allocated and unstructured branches. This should not be a
problem, as lowering a single instruction at any of the progressive
levels of abstraction provided by the RISC-V dialect down to regular 
assembly should not be difficult.

@@

The core contribution will be proof automation. By treating
instruction selection as a sequence of rewrites, we merely need to
verify that each of these rewrites individually preserves semantics.

This seems like much the same problem that we tackled in the previous
chapters, but there are some subtleties. For one, the rewrites
themselves should be simpler, with many rewrites having only a single
instruction as source and target. However, a Sail-derived semantics
model will be much larger than the handwritten model we'll be using for
the compiler IR, necessitating many more reasoning steps to simplify the
semantics for a single instruction. Many of the details that make the
Sail model so large might be irrelevant in our context, so I'll
investigate ways to simplify the model under specific assumptions. For
example, we could assume we're only compiling user-level programs, which might
allow us to simplify away any complexity to do with kernel-level
instructions in the model.

@@

During my internship at Amazon, I developed a symbolic simulator for a
simplified fragment of Arm assembly programs, using a handwritten model
of the ARM ISA semantics. My internship focused primarily on making this
symbolic simulator scale to being able to simulate hundreds of
instructions within a reasonable timeframe. I'll be using this symbolic
simulator as a basis for the translation validation tool I aim to
develop in this project, adapting it to the slightly different design constraints in
this context.

@@

Islaris is a Coq development, which builds upon Isla and Iris separation logic to 
prove properties about assembly programs using the power and expressivity of a 
theorem prover~\cite{sammlerIslarisVerificationMachine2022}. Their strategy starts by using Isla to transform a program into a trace, and then proving properties about this trace.
To remove the need to trust the SMT solver, Sammler\etal{} also investigated
the use of automated translation validation, which certifies that the trace produced
by Isla agrees with the lowering of the Sail semantics to Coq. This approach 
worked well for RISC-V, but unfortunately did not scale to the much larger Arm model.

Although I'll focus specifically on RISC-V, I'll still choose to reimplement 
the symbolic simulator in Lean, rather than reusing Isla in the same way, 
in the hopes that this approach might prove more scalable.

@@

In previous chapters, we've glossed over how, exactly, I'll give
semantics to the side-effectful version of our compiler IR. Zakowski et
al.~have written about the benefits of an interaction tree (ITree)
based semantics~\cite{zakowskiModularCompositionalExecutable2021, xiaInteractionTreesRepresenting2020}, so I'd like to use
ITrees also. Sadly, ITrees are coinductive, to properly capture infinite loops, 
and Lean does not support coinductive types.
This is why I'll initially be building a semantics for a fragment of the IR with
trivially terminating for-loops. 
In the current chapter I'll address this shortcoming, by building a Lean library for ITrees and using it to define semantics for potentially infinite loops,
answering research question \textsc{RQ3} in the process.

@@

Luckily, we can encode coinductives in Lean's existing logic, using
so-called \emph{quotients of polynomial functors} (QPFs). Avigad\etal{} have written about the low-level construction~\cite{avigadDataTypesQuotients2019}, 
upon which I expanded in my Master's thesis by prototyping a higher-level framework for defining
coinductive types, called QPFTypes. This framework is far from perfect,
and I'll spend some time fixing bugs in the framework. Still, the main
goal will be simply to define interaction trees; not to perfectly polish
the framework.

@@

The QPFTypes framework currently only has facilities for defining types;
to define corecursive functions that operate over ITrees we only have
the low-level corecursion principle to work with. However, in their
original development of ITrees, Xia\etal{} in fact provide a set of three
combinators which they claim make working with ITrees more tractable,
and remove the need for users to work directly with Coq's implementation
of corecursive function (which they claim is not compositional). As it
turns out, these combinators are rather similar to the corecursion
principle we do have available. So, for our ITree library
in Lean, we don't have to worry about general corecursive functions; it
suffices to implement the three basic combinators that Xia\etal{} provide.

@@

Isabelle, in particular, is relevant because its construction of
coinductive types, in terms of \emph{bounded natural functors} is
closely related to our construction in terms of QPFs~\cite{furerQuotientsBoundedNatural2022}. There are still new
challenges, however, because Lean's dependent types mean we have to
worry about the distinction between definitional equality (which is a
decidable, but weaker notion of equality) and propositional equality
(which captures more things that are intuitively equal, but requires
proof). Isabelle, in contrast, uses a weaker logical system where all
equalities are decidable.

@@

As mentioned, I'll work on the instruction selection project last.
I'll list the 3 chapters in order that I'll work on them, which differs from the
chapter order presented so far.

@@

\subsection*{Automatically Verified Instruction Selection using
Authoritative ISA Models}\label{scaling-high-assurance-translation-validation-to-authoritative-isa-models-1}
